Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-08-12 2024/OPTIMA/OPTIMA_2024_MCO.tex
add A.G. suggesions

%
\begin{abstract}
Solving a multicriteria optimization problem requires finding a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for approximate solving ``black-box'' multiobjective problems with multiextremal criteria. The proposed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set approximation. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriteria optimization. The effectiveness of different versions of the proposed algorithm is estimated on the base of a representative computational experiment.

@@

As it was mentioned above, the partial criteria contradictoriness leads to the necessity to introduce the set of non-dominated points called Pareto set as the general solution of the multicriteria problem \cite{Miettinen1999,Ehrgott2005,Pardalos2017}. Hereinafter, any point of the Pareto set will be called a \textit{partial solution} or just an \textit{efficient point}.

@@

One of the ways how this can be done is connected with the modification of the starting iterations of GSA. Namely, if $P \geq 1$ problems (\ref{eq:07}) have been already solved and the matrix (\ref{eq:15}) has been formed, then as initial trial coordinates, the points $x^q$, $0 \leq q \leq K$, can be used with values $z^q = F(\lambda,y^q)$. Moreover, the computation of the convolution (\ref{eq:05}) does not require the new calculation of criteria values at these points, because these values are taken from the matrix (\ref{eq:15}). After that, one can start optimization algorithm according to the Rules 1-5. Such the scheme reduces significantly the number of criteria computations when jointly solving the family of problems (\ref{eq:04}) with convolution (\ref{eq:05}) for different weight coefficients $\lambda$ \cite{ML_MCO_2023,Gergel2018,GergelKozinov2020}.

The second approach also aims to further decreasing the number of trials required to find the Pareto set. It suggests to change the form of the characteristics (\ref{eq:11}) in the GSA scheme. It is proposed to use the characteristic 
\begin{equation}
    \label{eq:16}
    R(i) = R_{gsa} (i) +  \alpha R_{ps} (i)
\end{equation}
instead of the characteristic (\ref{eq:11}), where $R_{gsa}(i)$ coincides with the expression (\ref{eq:11}) and the additional term $R_{ps}(i)$ depends on closeness of the points $y(x_{i-1})$, $y(x_{i})$ to the Pareto set, namely, the closer these points to the Pareto set, the greater the value $R_{ps}(i)$.

@@

After the construction of the separating hyperplane, for each trial point the $d_i$ distance is calculated and normalized according to the following expression:
\begin{equation}
    \label{eq:17}
d'_i=
\begin{cases}
  \begin{matrix}
     d_i / d_{max}, & d_i > 0 \\
     -d_i / d_{min}, & d_i < 0 
  \end{matrix}
\end{cases}, 
0 \leq i \leq k-1,
\end{equation}
where $d_i$ is the distance from the $i$-th point to the hyperplane, $d_{max}$, $d_{min}$ are maximum and minimum distances to the hyperplane.
The value $R_{ps}(i)$ from (\ref{eq:16}) for the interval $(x_{i-1},x_{i})$ is calculated as the sum of normalized distances (\ref{eq:17}) to $i$-th and $(i-1)$-th points. Let's denote this version of GSA as ML\_MGSA\_Dist.

@@

Solving several series of MCO problems with complex multiextremal criteria has demonstrated that the proposed approach to evaluating the set of efficient solutions is promising. Further research aims to develop the proposed approach by consideration of other machine learning models (in particular, artificial neural networks) to better classify trials and approximate the Pareto set.

>>>>

%
\begin{abstract}
Solving a multicriteria optimization problem requires finding a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for approximate solving ``black-box'' multiobjective problems with multiextremal criteria. The proposed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set approximation. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriteria optimization. Based on a representative computational experiment, the effectiveness of different versions of the proposed algorithm is estimated.

@@

As it was mentioned above, the contradictory nature of the subcriteria makes it necessary to introduce the set of non-dominated points called Pareto set as the general solution of the multicriteria problem \cite{Miettinen1999,Ehrgott2005,Pardalos2017}. Hereinafter, any point of the Pareto set will be called a \textit{partial solution} or just an \textit{efficient point}.

@@

One way of achieving this is to modify the starting iterations of GSA. Specifically, if $P \geq 1$ problems (\ref{eq:07}) have already been solved and the matrix (\ref{eq:15}) has been formed, then the points $x^q$, $0 \leq q \leq K$, with values $z^q = F(\lambda,y^q)$ can be used as initial trial coordinates. Moreover, the computation of the convolution (\ref{eq:05}) does not require the recalculation of criteria values at these points, since these values are taken from the matrix (\ref{eq:15}). The optimization algorithm can then be started according to the Rules 1-5. This scheme significantly reduces the number of criteria computations when solving simultaneously the family of problems (\ref{eq:04}) with convolution (\ref{eq:05}) for different weight coefficients $\lambda$ \cite{ML_MCO_2023,Gergel2018,GergelKozinov2020}.

The second approach also aims to further reduce the number of trials required to find the Pareto set. It suggests to change the form of the characteristics (\ref{eq:11}) in the GSA scheme. It is proposed to use the characteristic 
\begin{equation}
    \label{eq:16}
    R(i) = R_{gsa} (i) +  \alpha R_{ps} (i)
\end{equation}
instead of the characteristic (\ref{eq:11}), where $R_{gsa}(i)$ coincides with the expression (\ref{eq:11}) and the additional term $R_{ps}(i)$ depends on closeness of the points $y(x_{i-1})$, $y(x_{i})$ to the Pareto set, namely, the closer these points to the Pareto set, the greater the value $R_{ps}(i)$.

@@

After constructing the separating hyperplane, for each trial point the $d_i$ distance is calculated and normalized according to the following expression:
\begin{equation}
    \label{eq:17}
d'_i=
\begin{cases}
  \begin{matrix}
     d_i / d_{max}, & d_i > 0 \\
     -d_i / d_{min}, & d_i < 0 
  \end{matrix}
\end{cases}, 
0 \leq i \leq k-1,
\end{equation}
where $d_i$ is the distance from the $i$-th point to the hyperplane, $d_{max}$, $d_{min}$ are maximum and minimum distances to the hyperplane.
The value $R_{ps}(i)$ from (\ref{eq:16}) for the interval $(x_{i-1},x_{i})$ is calculated as the sum of normalized distances (\ref{eq:17}) to $i$-th and $(i-1)$-th points. Let's denote this version of GSA as ML\_MGSA\_Dist.

@@

Solving several series of MCO problems with complex multiextremal criteria has demonstrated the potential of the proposed approach to evaluate the set of efficient solutions. Further research aims at extending the proposed approach by considering other machine learning models (in particular, artificial neural networks) to better classify the trials and approximate the Pareto set.
